Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-04 appendix.tex
hopefully final resolution to sec 5.3, major revision, please double check (esp. the new appendix D3)

\subsection{Stability conditions in the low-frequency limit}
In the low-frequency limit we assume $|\hat{\upsilon}|\ll 1$ and
approximate Eq. \ref{iso_ode0} as 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \ii h q \hat{k}
    f(\zhat)\right]W^\prime +
  \hat{\upsilon}^2\left(1+\hat{k}^2\right)W. \label{iso_ode1}
\end{align}
\subsubsection{Necessity of vertical shear for
  instability}\label{integral_relation} 
We can establish a necessary condition for instability by multiplying
Eq. \ref{iso_ode1} by $\rho W^*$ and integrate vertically from
$\zhat=\zhat_1$ to $z=\zhat_2$. We neglect boundary 
terms when integrating by parts, by assuming $W$ or
$W^\prime$ vanishes at the boundaries, or that the boundaries are 
sufficiently far away so that the boundary terms are negligible because of the
decaying background density with increasing height. Then,
\begin{align}
  \hat{\upsilon}^2\left(1+\hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|W|^2d\zhat
  =\int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\zhat 
  +\ii h q \hat{k}\int_{\zhat_1}^{\zhat_2}\rho f(\zhat) W^*W^\prime d\zhat.\label{integral_relation1}
\end{align}
It follows that for instability ($\imag\hat{\upsilon}>0$), it is necessary to
have $q\neq0$ or more generally $d\Omega^2/dz\neq 0$, i.e. there must
be vertical shear. A similar conclusion can be reached in the
equivalent linear problem in global cylindrical disk geometry.

\subsubsection{Maximum growth rate}\label{max_growth1}
Here we show that the growth rate is limited by the maximum vertical
shear rate in the domain. % Writing $\hat{\upsilon} = \hat{\omega} +
% \ii\hat{\upsilon}$ with real $\hat{\omega}$ and $\hat{\upsilon}$, 
The real and imaginary parts of 
Eq. \ref{integral_relation1} are
\begin{align}
  \left(\hat{\omega}^2-\hat{\sigma}^2\right)\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
  \int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\hat{z}
  \notag
  =\real\left[\ii
    h\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\zhat\right], \\
   2\hat{\omega}\hat{\sigma}\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z}
  =\imag\left[\ii
    h\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right],
\end{align}
where we recall $\hat{\omega} = \real \hat{\upsilon}$ and
$\hat{\sigma}=\imag\hat{\upsilon}$. 
Adding the square of these equations give
\begin{align}
  \left[|\hat{\upsilon}|^2\left(1+\hat{k}^2\right)
    \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
    \int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}\right]^2
  +4\hat{\sigma}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}
  =\left|\ii
    h\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
It is clear that
\begin{align}\label{sigma_finite_domain} 
  4\hat{\sigma}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime
  \right|^2d\hat{z} 
  \leq\left|
    h\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
On the left hand side of this inequality, we apply the Cauchy-Schwarz
inequality to obtain
\begin{align}
  \left( \int_{\zhat_1}^{\zhat_2}\rho
    |W|\left|W^\prime \right|d\hat{z}\right)^2\leq
  \int_{\zhat_1}^{\zhat_2}\rho 
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}.
\end{align}
On the right hand side of Eq. \ref{sigma_finite_domain} we have
\begin{align}
  \left|\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|\leq \int_{\zhat_1}^{\zhat_2}\rho
  \left|f(\hat{z})W^*W^\prime \right|d\hat{z}
  \leq
  \mathrm{max}\left(|f|\right)\int_{\zhat_1}^{\zhat_2}\rho
  |W|\left|W^\prime \right|d\hat{z},
\end{align}
where $\mathrm{max}(|f|)$ is the maximum value of $|f|$ in
$\zhat\in[\zhat_1,\zhat_2]$. Inserting these inequalities into
Eq. \ref{sigma_finite_domain} gives
\begin{align}\label{max_growth}
  |\hat{\sigma}|\leq
  \frac{h |\hat{k} q|}{2\sqrt{1+\hat{k}^2}}\mathrm{max}(|f|) < \frac{h |q|}{2}\mathrm{max}(|f|) . 
\end{align}
%Recalling that $f$ represents vertical shear (Eq. \ref{fz_shear}), 
It
follows that the maximum possible growth rate of unstable modes,
satisfying the above boundary conditions, is limited by the maximum
vertical shear rate in the domain considered,
\begin{align}
  \sigma < \mathrm{max} \left|r\frac{d\Omega}{dz}\right|, 
\end{align}
as expected on physical grounds.

@@

% For vertical domains smaller 
% than this height, the growth rate is limited by $|f|$ at the vertical
% boundaries considered. For $|h\hat{z}|\ll1$ we have $f\simeq
% \hat{z}$, in which case we can expect growth rates to increase at most linearly
% with respect to the vertical domain size.   %no faster than linear

@@

\subsection{Maximum critical cooling time}\label{max_cool}
 % {\bf [at the moment for reference
 %  only, not to be included in final version]}\\
Consider the simplified dispersion relations for marginal stability,
Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2}. Writing 
$X= \hat{\omega}\khat$, $\theta = (hq\khat)^2\leq1$ and  treating $M$ as a
continuous variable, we find from Eq. \ref{relax_cond_simp1} that 
%\begin{align}
%  &x^4  = x^2 + M(M+1)(1-\theta), \label{simp1}\\
%  &x^2  = \frac{\gamma-1}{h q}\beta_c (1+2M)^2x - 2M(M+1)\label{simp2},
%\end{align}
%where and
\begin{align}
  2 X \frac{d X}{dM} = \frac{X^2(1-\theta)(2M+1)}{X^2 +
  2M(M+1)(1-\theta)}, 
\end{align}
and from Eq. \ref{relax_cond_simp2} that
\begin{align}
  2X\frac{dX}{dM} = \left[X^2 +
  2M(M+1)\right]\left(\frac{d\ln{\beta_c}}{dM} + \frac{4}{2M+1} +
  \frac{d\ln{X}}{dM}\right) - 2(2M+1). 
\end{align}
We eliminate $dX/dM$, making use of Eq. \ref{relax_cond_simp1} in the process, to obtain 
\begin{align}
  (2M+1)\frac{d\ln{\beta_c}}{dM} = \frac{\mathcal{N}}{\mathcal{D}}, 
\end{align}
where the denominator $\mathcal{D}$ is
\begin{align}
\mathcal{D} = 2 \left[X^2 +
  2M(M+1)(1-\theta)\right]\left[X^2 + 2M(M+1)\right]>0, 
\end{align}
and the numerator $\mathcal{N}$ is 
\begin{align}
  \mathcal{N} = -(1-\theta)\left[ 2M(M+1)(2M+1)^2 + 12X^2M(M+1)\right]
  - (3+\theta)X^2 < 0.
\end{align}
Hence,
\begin{align}
  \frac{d\beta_c}{dM}<0, 
\end{align}
for all $M$ provided that $\theta\leq 1$. This implies 
$\mathrm{max}(\beta_c)$ occurs at $M=0$.

>>>>

\subsection{Maximum growth rate  in the low-frequency limit}
Here we consider the low-frequency limit $|\hat{\upsilon}|\ll 1$ and 
show that the growth rate is limited by the maximum vertical
shear rate in the domain. 
%In the low-frequency limit we assume $|\hat{\upsilon}|\ll 1$ and
We approximate Eq. \ref{iso_ode0} as 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \ii h q \hat{k}
    f(\zhat)\right]W^\prime +
  \hat{\upsilon}^2\left(1+\hat{k}^2\right)W. \label{iso_ode1}
\end{align}
%\subsubsection{Necessity of vertical shear for
%  instability}\label{integral_relation} 
%We can establish a necessary condition for instability 
We multiply Eq. \ref{iso_ode1} by $\rho W^*$ and integrate vertically from
$\zhat=\zhat_1$ to $z=\zhat_2$. We neglect boundary 
terms when integrating by parts, by assuming $W$ or
$W^\prime$ vanishes at the boundaries, or that the boundaries are 
sufficiently far away so that the boundary terms are negligible because of the
decaying background density with increasing height. Then,
\begin{align}
  \hat{\upsilon}^2\left(1+\hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|W|^2d\zhat
  =\int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\zhat 
  +\ii h q \hat{k}\int_{\zhat_1}^{\zhat_2}\rho f(\zhat) W^*W^\prime d\zhat.\label{integral_relation1}
\end{align}
It follows that for instability ($\imag\hat{\upsilon}>0$), it is necessary to
have $q\neq0$ or more generally $d\Omega/dz\neq 0$, i.e. there must
be vertical shear. 
%A similar conclusion can be reached in the
%equivalent linear problem in global cylindrical disk geometry.  
%\subsubsection{Maximum growth rate}\label{max_growth1}

The real and imaginary parts of 
Eq. \ref{integral_relation1} are
\begin{align}
  \left(\hat{\omega}^2-\hat{\sigma}^2\right)\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
  \int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\hat{z}
  \notag
  =\real\left[\ii
    h\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\zhat\right], \\
   2\hat{\omega}\hat{\sigma}\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z}
  =\imag\left[\ii
    h\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right],
\end{align}
where we recall $\hat{\omega} = \real \hat{\upsilon}$ and
$\hat{\sigma}=\imag\hat{\upsilon}$. 
Adding the square of these equations give
\begin{align}
  \left[|\hat{\upsilon}|^2\left(1+\hat{k}^2\right)
    \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
    \int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}\right]^2
  +4\hat{\sigma}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}
  =\left|\ii
    h\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
It is clear that
\begin{align}\label{sigma_finite_domain} 
  4\hat{\sigma}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime
  \right|^2d\hat{z} 
  \leq\left|
    h\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
On the left hand side of this inequality, we apply the Cauchy-Schwarz
inequality to obtain
\begin{align}
  \left( \int_{\zhat_1}^{\zhat_2}\rho
    |W|\left|W^\prime \right|d\hat{z}\right)^2\leq
  \int_{\zhat_1}^{\zhat_2}\rho 
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}.
\end{align}
On the right hand side of Eq. \ref{sigma_finite_domain} we have
\begin{align}
  \left|\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|\leq \int_{\zhat_1}^{\zhat_2}\rho
  \left|f(\hat{z})W^*W^\prime \right|d\hat{z}
  \leq
  \mathrm{max}\left(|f|\right)\int_{\zhat_1}^{\zhat_2}\rho
  |W|\left|W^\prime \right|d\hat{z},
\end{align}
where $\mathrm{max}(|f|)$ is the maximum value of $|f|$ in
$\zhat\in[\zhat_1,\zhat_2]$. Inserting these inequalities into
Eq. \ref{sigma_finite_domain} gives
\begin{align}\label{max_growth}
  |\hat{\sigma}|\leq
  \frac{h |\hat{k} q|}{2\sqrt{1+\hat{k}^2}}\mathrm{max}(|f|) < \frac{h |q|}{2}\mathrm{max}(|f|) . 
\end{align}
%Recalling that $f$ represents vertical shear (Eq. \ref{fz_shear}), 
It
follows that the maximum possible growth rate of unstable modes,
satisfying the above boundary conditions, is limited by the maximum
vertical shear rate in the domain considered,
\begin{align}
  |\sigma| < \mathrm{max} \left|r\frac{d\Omega}{dz}\right|, 
\end{align}
as expected on physical grounds. %A similar conclusion can be reached in the
%equivalent linear problem in global cylindrical disk geometry.

@@

@@

\subsection{Maximum critical cooling time}\label{max_cool}
 % {\bf [at the moment for reference
 %  only, not to be included in final version]}\\
Consider the simplified dispersion relations for marginal stability,
Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2}. Writing 
$X= \hat{\omega}\khat$, $\theta = (hq\khat)^2\leq1$ and  treating $M$ as a
continuous variable, we find from Eq. \ref{relax_cond_simp1} that 
%\begin{align}
%  &x^4  = x^2 + M(M+1)(1-\theta), \label{simp1}\\
%  &x^2  = \frac{\gamma-1}{h q}\beta_c (1+2M)^2x - 2M(M+1)\label{simp2},
%\end{align}
%where and
\begin{align}
  2 X \frac{d X}{dM} = \frac{X^2(1-\theta)(2M+1)}{X^2 +
  2M(M+1)(1-\theta)}, 
\end{align}
and from Eq. \ref{relax_cond_simp2} that
\begin{align}
  2X\frac{dX}{dM} = \left[X^2 +
  2M(M+1)\right]\left(\frac{d\ln{\beta_c}}{dM} + \frac{4}{2M+1} +
  \frac{d\ln{X}}{dM}\right) - 2(2M+1). 
\end{align}
We eliminate $dX/dM$, making use of Eq. \ref{relax_cond_simp1} in the process, to obtain 
\begin{align}
  (2M+1)\frac{d\ln{\beta_c}}{dM} %= \frac{\mathcal{N}}{\mathcal{D}}, 
  = -\frac{(1-\theta)M(M+1)\left[ 2(2M+1)^2 + 12X^2\right]
  + (3+\theta)X^2}{2 \left[X^2 +
  2M(M+1)(1-\theta)\right]\left[X^2 + 2M(M+1)\right]}.  
\end{align}
%where the denominator $\mathcal{D}$ is
%\begin{align}
%\mathcal{D} = 2 \left[X^2 +
 % 2M(M+1)(1-\theta)\right]\left[X^2 + 2M(M+1)\right]>0, 
%\end{align}
%and the numerator $\mathcal{N}$ is 
%\begin{align}
 % \mathcal{N} = -(1-\theta)\left[ 2M(M+1)(2M+1)^2 + 12X^2M(M+1)\right]
 % - (3+\theta)X^2 < 0.
%\end{align}
Hence,
\begin{align}
  \frac{d\beta_c}{dM}<0, 
\end{align}
for all $M\geq 0$ if $\theta\leq 1$. This implies 
$\mathrm{max}(\beta_c)$ occurs at $M=0$.
